---
title: "Inequality as a Predictor for Social Instability"
author: "JC De Orellana Sanchez"
date: "`r Sys.Date()`"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}

library(tidyverse)
library(modelsummary)   
library(estimatr)       
library(plm)            
library(mfx)            
library(infer)
library(corrplot)
library(ggrepel)
library(ggdag)
library(patchwork)

```
## Introduction

History, sociology, and other social sciences have cautiously avoided or emphatically rejected predicting or forecasting social unrest. Their goal, these disciplines maintain, is to explain phenomena through empirical observation, testing, and theorizing. Data science, political economy, and public policy, on the other hand, seek to define variables an methods to measure and predict the costs and effects of policies as well as the domestic and external factors that influence and trigger dynamics societies and the economy. This paper is the first step in a long term project that seeks to bridge both perspectives to understand what economic, political, social, environmental, and international factors may lead to social instability (discontent, unrest, and violence). Specifically, the paper asks whether inequality--economic and social--are good predictors of social instability.

-   **answer**

Relying on data collected from large time-series datasets containing economic, political, and social social data collected for countries around the globe, the results of this study are still inconclusive. It posits that predicting social unrest and instability seems possible to an extent, but greater attention needs to be paid to the variables and metrics used, and locally specific circumstances. The answer to the first, broader question is partial. Rather than determining the factors that if properly accounted might lead to social unrest and instability, at this stage the study can only show which ones unperformed and do not appear as good indicators. The second question also returned mixed results. Some inequality variables are better at predicting social instability, at least in the terms defined in this paper: based on political terror and the presence of riots and protests after elections. The overall answer answer to both questions seems to be that it is necessary to revisit the variable and metrics with which political instability is defined. It appears crucial to find metrics that account for social instability that are not as extreme as terrorism or actual violence. Perhaps insecurity measured in terms of robbery, theft, assault, and organized crime, as well as the frequency of protests, strikes, and maybe engagement with support to "alternative" media and political candidates.

For centuries, scientists have worked to develop methods, tools, and metrics to forecast the weather and earthquakes. Studying the atmosphere, weather, and climate, meteorologists use scientific principles to understand and predict weather patterns and atmospheric phenomena. As of today, they have refined their methods to the point of being able to do short-term forecasts (up to 48 hours) with a high degree of precission. However, the accuracy diminishes depending on how far in the future they try to predict the weather. 3 - 7 days forecasts are still accurate but their precision in terms of details like temperatures or precipitation dimish. 8 - 14 days predictions are even less accurate as they provide an idea of what could happen based on historical data and the observation of occuring trends. Extended-range forecasts of more than two weeks are mostly experimental and focus on overall climate patterns and anomalies rather than on predicting specific weather conditions.

To reach the present degrees of accuracy for the prediction and forecasting of weather conditions, meteorology relies on data produced by systematic observation, measurement, and recording of climate phenomena. Recording weather conditions was only possible thanks to the invention and use of tools such as the thermometer, barometer, and hygrometer in the 17th century, as they allowed the quantification and scaling of temperature, wind speed, and humidity. These instruments were later followed by radiosondes, weather balloons, and radars. The incorporation of models based on fluids dynamics and thermodynamics in the 20th century turned the possibility to understand and predict the weather into reality. The incorporation of satellite imagery and development of numerical weather prediction thanks to powerful computers, has allowed to factor in the extant data and methods to develop simulation models that have brought meteorology to its current state of affairs.

Similarly, seismology is today capable of predicting earthquakes within a few seconds so as to alert the population via instant message systems so they can act quickly and find safe spaces.

The forecasting capabilities of meteorology and seismology rely on measurement tools, mathematical and physics models, the attention to historical trends, and the incorporation of computational techniques. The main point of this paper is that the prediction of social unrest and instability can be achieved by uncovering the right social political, and economic dynamics to focus in, understanding the impact of the environment in them, and factoring in specific cultural markers, as well as historical and international trends. To do so, social and data scientists must ascertain what are the correct phenomena to measure (like temperature and pressure with the weather), create the best "sensors" to capture their variation, and develop adequate mathematical and computational models to explain and forecast them. 

Like with seismology, perhaps forecasting the precise moment when a riot, rebellion, or revolution might arise will only be achievable for the short term, based on the constant updating of new data and intelligence. Yet, like with meteorology, the continuous trace of certain dynamics, patterns, phenomena, and history may help predict for the mid and long term that, if not corrections are incorporated, the rise of discontent and the likelihood for it to derive into instability. 

### Literature Review

The paper follows the work of Peter Turchin, Doug McAdam, Sidney Tarrow, Charles Tilly. Turchin is the proponent of an analytic and predictive science of history he calls Cliodynamics. According to him, there may be general regularities characterizing the dynamics of human societies, and it is because social systems are extremely complex that we need mathematical models to grasp and formulate them properly (Turchin, 2011, 173). Like scientific prediction, historical prediction can be used to test theories. "In a scientific prediction exercise we want to use the degree of match between observables to infer the validity of a theory," and since "no theory makes perfect predictions, we ... want to compare the match between predictions and data for two (or more) theories." For history, Turchin argues, one can use two or more theories and see how well they predict phenomena in the past, thus making retrospective predictions or "retrodictions" (Turchin, 2011, 169-170).

The paper also follows Turchin's work on political instability and Doug McAdam', Sidney Tarrow', Charles Tilly's publications on contentious politics. For Turchin, under the structural-demographic theory, human societies are systems with three main compartments: the general population, the elite, and the state (Turchin, 2013) . These interact with each other and with socio-political instability built in via a web of nonlinear feedbacks (Figure 1). For him social pressure can be quatified for instability using three variables, the political stress indicator, the dynamic of real wages, and elite dynamics. Complementing Turchin's model/theory, the paper also builds upon Doug McAdam', Sidney Tarrow', Charles Tilly's methods for measuring mechanisms of contention (McAdam, Tarrow & Tilly, 2008). According to the authors, the best way to explain and predict contentious politics is by focusing on causality. They argue that the best way to find the cause(s) of contention is by analyzing mechanisms and processes.

## Data

### Datasets

-   Teorell, Jan, Aksel Sundström, Sören Holmberg, Bo Rothstein, Natalia Alvarado Pachon, Cem Mert Dalli, Rafael Lopez Valverde & Paula Nilsson. 2024. The Quality of Government Standard Dataset, version Jan24. University of Gothenburg: The Quality of Government Institute, https://www.gu.se/en/quality-government doi:10.18157/qogstdjan24Quality of Government dataset. Time series released on January 2024

-   The Political Terror Scale dataset. https://www.politicalterrorscale.org/
    Amnesty International and State Department sets

-   United Nations Development Program (https://data.undp.org/access-all-data). Unemployment and Informal Employment data. 

-   WorldBank. Informal Economy Database. Elgin, C., M. A. Kose, F. Ohnsorge, and S. Yu. 2021. “Understanding Informality.” CERP Discussion Paper 16497, Centre for Economic Policy Research, London. https://www.worldbank.org/en/research/brief/informal-economy-database


### **Variables and Metrics (how are variables measured)**

-   **summary statistics table**

-   **non-measured but important**

## Method/s

-   why the method

-   **causality**

```{r, echo = FALSE}
# DRAWING DAGs -------------------------------------

# SI: Social Instability
# Is: Social Inequality
# Ie: Economic Inequality

theme_set(theme_void() + theme(legend.position = 'none'))
gh <- dagify(SI ~ Ie + IS) |>
  ggdag_parents("SI") +
  labs(title = "General Hipothesis")

(gh)

```

```{r,  echo=FALSE}
# DRAWING DAGs -------------------------------------

# SI: Social Instability
# Ie: Economic Inequality
# C: Corruption
# U: Unemployment + Precarious/Informal Employment
# BS: Basic Services

theme_set(theme_void() + theme(legend.position = 'none'))
h1 <- dagify(SI ~ Ie + C + U + BS, Ie ~ C + U + BS) |> 
  ggdag_parents("SI") + 
  labs(title = "Theory 1") 

h2 <- dagify(SI ~ Ie + C + U + BS, C  ~ Ie, U ~ Ie, BS ~ Ie) |> 
  ggdag_parents("SI") + 
  labs(title = "Theory 2") 

(h1 + h2)
```

```{r,  echo=FALSE}
# DRAWING DAGs -------------------------------------

# SI: Social Instability
# Is: Social Inequality
# Ie: Economic Inequality
# C: Corruption
# U: Unemployment + Precarious/Informal Employment
# BS: Basic Services
# T: Trust
# E: Education

theme_set(theme_void() + theme(legend.position = 'none'))
h3 <- dagify(SI ~ Is + C + U + BS + T, Is ~ C + U + BS + E + Ie, T ~ Is, Ie ~ C + U + BS + E) |>
  ggdag_parents("SI") +
  labs(title = "Theory 3")

h4 <- dagify(SI ~ Is + C + U + BS + T, Is ~ C + U + BS + E + Ie, C ~ T, Ie ~ C + U + BS + E) |>
  ggdag_parents("SI") +
  labs(title = "Theory 4")

(h3 + h4)

```


## Results

```{r, echo=FALSE}

# load data -------------------------------------------------------------------
source("data/instability_data.r")

# Social Instability Data Frame-------------------------------------------------
soc_ins <- instab_dataset |>
  filter(year %in% 1980:2022) |>
  dplyr::select(Country = cname, Year = year,
         # corruption
         poltcl_corrptn_idx = vdem_corr,
         media_corrupt = vdem_mecorrpt,
         election_vote_buying = vdem_elvotbuy,
         control_of_corruption_estimate = wbgi_cce,
         
         # trust
         trust_in_people = wvs_trust,
         
         # social instability. Check the ones that have the most data
         polit_stab_absence_of_viol_terror = wbgi_pve, #1996-2022
         Viol_before_elect = nelda_vcdbe, #1946-2020
         Riot_n_prot_after_elect = nelda_rpae, #1946-2020
         gd_ptsa, # Political Terror Scale Amnesty International 1976-2021
         gd_ptss, # Political Terror Scale State Department 1976-2021
         
         #education
         sch_enrlmnt_prmry_gross = wdi_gerp, # "School enrollment, primary (% gross)"
         ed_attnmnt_fem_15_24 = gea_ea1524f, #"Educational Attainment (15-24 years, Female)"
         ed_attnmnt_mal_15_24 = gea_ea1524m, #"Educational Attainment (15-24 years, Male)"
         
         # unemploy
         wdi_unempilo,     # Unemployment, total (% of total labor force) (modeled ILO)
         wdi_lfpilo15,     # Labor force participation rate (% of total ages 15+) (modeled ILO)
         wdi_unempmilo,    # total male unemp
         wdi_unempyilo,    # total youth unemployment (15-24)
         wdi_unempymilo,   # total male youth unemployment (15-24)
         # informal and self employment
         SEMP_p,
         Inf_Emp_Agr_Total_Thosnds,
         Inf_Emp_Non_Agr_Total_Thosnds,
         Inf_Emp_Employess_Total_Thosnds,
         Inf_Emp_Self_Emp_Total_Thosnds,

         # basic services
         gov_size  = fi_sog, #"Size of Government: Expenditures, Taxes and Enterprises"
         wdi_acelu, wdi_acelr,   # electricity urban rural
         wdi_chexppgdp,          # health expenditure
         wdi_hwf,                # hand washing
         who_sanittot, # Total population using basic sanitation services (%)
         ihme_hle_0104t, # Healthy life years 1-4 years total
         
         # inflation
         wdi_inflation, #Inflation, consumer prices
         
         wdi_popgr, #Population growth (annual %)
         
         # econ ineq (unequal distribution of income and wealth within a society)
         #wdi_povgap215, #only 1991 observations
         income_share_high_10pct = wdi_incsh10h,
         income_share_low_10pct = wdi_incsh10l,
         wdi_gdpcapgr, #GDP per capita growth (annual %) 
         Gini_Index,
         Income_inequality,
         
         # social ineq
         egal_comp_idx = vdem_egal, # "Egalitarian component index"
         sch_enrlmnt_tertiary_gross = wdi_gert,
         Human_Inequality_Coefficient,
         human_cap_idx = egov_hci
  )|>
  mutate(pol_terr_scale_avg = if_else(!is.na(gd_ptsa) & !is.na(gd_ptss),
                                      (gd_ptsa + gd_ptss) / 2, NA_real_))

```

## Theory 1. Economic instability can predict social instability

### Correlation Plot

```{r, echo=FALSE}
soc_ins |> 
  dplyr::select(Country, "Political Terror" = pol_terr_scale_avg, 
                "GDP/cap growth(%)" = wdi_gdpcapgr, 
                "Inc Share High 10%" = income_share_high_10pct, 
                "Inc Share Low 10%" = income_share_low_10pct, Gini_Index,
                Income_inequality, 
                "Young Unemp Male" = wdi_unempymilo, 
                "Corruption" = control_of_corruption_estimate, "Government Size" = gov_size) |>
  dplyr::select_if(is.numeric) |> 
  cor(use = "pairwise.complete.obs") |> 
  corrplot(method = "number", type = 'lower')
```

```{r, echo=FALSE}
soc_ins |> 
  dplyr::select(Country, "Political Terror" = pol_terr_scale_avg, 
                "GDP/cap growth(%)" = wdi_gdpcapgr, 
                "Inc Share High 10%" = income_share_high_10pct, 
                "Inc Share Low 10%" = income_share_low_10pct, Gini_Index,
                Income_inequality, "Young Unemp Male" = wdi_unempymilo, 
                "Corruption" = control_of_corruption_estimate, "Government Size" = gov_size) |>
  dplyr::select_if(is.numeric) |> 
  cor(use = "pairwise.complete.obs") |>
  corrplot(type = 'upper', method = "shade") 

```

### Scatter Plots

```{r, echo=FALSE}

# Theory 1 ---------------------------------------------------------------------
## Economic Inequality on Social Instability ------------
### plotting the linear regression for income inequality ----
soc_ins |> 
  filter(Year %in% 2000:2022) |>
  ggplot(aes(x = Income_inequality,
             y = pol_terr_scale_avg
             )) +
  geom_jitter() +
  geom_smooth(method = lm) +
  labs(title = "Economic Inequality on Social Instability (2000-2022)", 
       subtitle = "Political Terror ~ Income Inequality", x = "Political Terror", 
       y = "Economic Inequality") +
  theme_minimal() +
  theme(
    axis.title = element_text(color = "darkgrey"),
    legend.position = "bottom",  
    axis.text.x = element_text(size = 7),
    axis.text.y = element_text(size = 7)
  )

soc_ins |> 
  filter(Year %in% 2000:2022) |>
  ggplot(aes(x = wdi_gdpcapgr,
             y = pol_terr_scale_avg
  )) +
  geom_jitter() +
  geom_smooth(method = lm) +
  labs(title = "Economic Inequality on Social Instability (2000-2022)", 
       subtitle = "Political Terror ~ GDP/Capita Growth(%)", x = "Political Terror", 
       y = "GDP/Capita Growth(%)") +
  theme_minimal() +
  theme(
    axis.title = element_text(color = "darkgrey"),
    legend.position = "bottom",  
    axis.text.x = element_text(size = 7),
    axis.text.y = element_text(size = 7)
  )

```

### Linear Regression Economic Enequality Indicators on Social Instability (Political Terror) with and without Controls

```{r, echo=FALSE}
### linear regression 

with(soc_ins, {
  list(
    "GDP/cap growth (%)" = lm_robust(formula = pol_terr_scale_avg ~
                                       wdi_gdpcapgr + factor(Year) + 
                                       factor(Country)),
    "GDP/cap growth (%) (Controls)" = lm_robust(formula = pol_terr_scale_avg ~
                                                  wdi_gdpcapgr + wdi_unempymilo + 
                                                  control_of_corruption_estimate + gov_size + 
                                                  factor(Year) + factor(Country)),
    "Income Share High 10%" = lm_robust(formula = pol_terr_scale_avg ~
                                  income_share_high_10pct + factor(Year) + factor(Country)),
    "Income Share High 10% (Controls)" = lm_robust(formula = pol_terr_scale_avg ~
                                   income_share_high_10pct + wdi_unempymilo + 
                                     control_of_corruption_estimate + gov_size + factor(Year) + 
                                   factor(Country)),
    "Income Share Low 10%" = lm_robust(formula = pol_terr_scale_avg ~
                                         income_share_low_10pct + factor(Year) + 
                                         factor(Country)),
    "Income Share Low 10% (Controls)" = lm_robust(formula = pol_terr_scale_avg ~
                                     income_share_low_10pct + gov_size + 
                                     wdi_unempymilo + control_of_corruption_estimate + 
                                     factor(Year) + factor(Country)),
    "Gini index" = lm_robust(formula = pol_terr_scale_avg ~
                             Gini_Index + factor(Year) + factor(Country)),
    "Gini index (Controls)"  = lm_robust(formula = pol_terr_scale_avg ~
                              Gini_Index + gov_size + wdi_unempymilo + 
                                control_of_corruption_estimate + factor(Year) + factor(Country)),
    "Income inequality" = lm_robust(formula = pol_terr_scale_avg ~
                                 Income_inequality + factor(Year) + factor(Country)), 
    
    "Income inequality (Controls)" = lm_robust(formula = pol_terr_scale_avg ~
                                 Income_inequality + gov_size + wdi_unempymilo + 
                                   control_of_corruption_estimate + factor(Year) + factor(Country))
    ) |> 
    msummary(coef_omit = "factor", stars = TRUE)
  })
```

### Linear Regression Economic Enequality Indicators on Social Instability (Riots/Protests after election) with and without Controls

```{r, echo=FALSE}
### linear regression for riots and protests after elections ----
with(soc_ins, {
  list(
    "GDP/cap growth (%)" = lm_robust(formula = Riot_n_prot_after_elect ~
                                       wdi_gdpcapgr + factor(Year) + 
                                       factor(Country)),
    "GDP/cap growth (%) (Controls)" = lm_robust(formula = Riot_n_prot_after_elect ~
                                                  wdi_gdpcapgr + wdi_unempymilo + 
                                                  control_of_corruption_estimate + gov_size + 
                                                  factor(Year) + factor(Country)),
    "Income Share High 10%" = lm_robust(formula = Riot_n_prot_after_elect ~
                                          income_share_high_10pct + factor(Year) + 
                                          factor(Country)),
    "Income Share High 10% (Controls)" = lm_robust(formula = Riot_n_prot_after_elect ~
                                          income_share_high_10pct + wdi_unempymilo + 
                                            control_of_corruption_estimate + gov_size + 
                                            factor(Year) + factor(Country)),
    "Income Share Low 10%" = lm_robust(formula = Riot_n_prot_after_elect ~
                                         income_share_low_10pct + factor(Year) + 
                                         factor(Country)),
    "Income Share Low 10% (Controls)" = lm_robust(formula = Riot_n_prot_after_elect ~
                                         income_share_low_10pct + wdi_unempymilo + 
                                           control_of_corruption_estimate + gov_size 
                                         + factor(Year) + factor(Country)),
    "Gini index" = lm_robust(formula = Riot_n_prot_after_elect ~
                               Gini_Index + factor(Year) + factor(Country)),
    "Gini index (Controls)" = lm_robust(formula = Riot_n_prot_after_elect ~
                               Gini_Index + wdi_unempymilo + 
                                 control_of_corruption_estimate + gov_size + 
                                 factor(Year) + factor(Country)),
    "Income inequality" = lm_robust(formula = Riot_n_prot_after_elect ~
                                      Income_inequality + factor(Year) + 
                                      factor(Country)),
    "Income inequality (Controls)" = lm_robust(formula = Riot_n_prot_after_elect ~
                                      Income_inequality + wdi_unempymilo + 
                                        control_of_corruption_estimate + gov_size 
                                      + factor(Year) + factor(Country))
    ) |> 
    msummary(coef_omit = "factor", stars = TRUE)
  })
```

### Corruption, Unemployment, and Basic Services on Economic Inequality

```{r, echo=FALSE}
## Corruption on Economic Inequality ------------
### plotting the linear regression
soc_ins |>
  filter(Year %in% 2000:2022) |>
  ggplot(aes(x = control_of_corruption_estimate,
             y = Income_inequality 
             )) +
  geom_jitter() +
  geom_smooth(method = lm) +
  labs(title = "Corruption on Economic Inequality (2000-2022)", 
       subtitle = "Income Inequality ~ Corruption", x = "Corruption", 
       y = "Income Inequality") +
  theme_minimal() +
  theme(
    axis.title = element_text(color = "darkgrey"),
    legend.position = "bottom",  
    axis.text.x = element_text(size = 7),
    axis.text.y = element_text(size = 7)
  )
```

```{r, echo=FALSE}
### linear regression 
with(soc_ins, {
  list("Corruption" = lm_robust(formula = Income_inequality ~ 
              control_of_corruption_estimate)) |>
         msummary(coef_omit = "factor", stars = TRUE)
})
```

```{r, echo=FALSE}
## Unemployment on Economic Inequality ------------
soc_ins |> 
  filter(Year %in% 2000:2022) |>
  ggplot(aes(x = wdi_unempyilo,
             y = Income_inequality)) +
  geom_point() +
  geom_smooth(method = lm) +
  labs(title = "Unemployment on Economic Inequality (2000-2022)", 
       subtitle = "Income Inequality ~ Unemployment", x = "Unemployment", 
       y = "Income Inequality") +
  theme_minimal() +
  theme(
    axis.title = element_text(color = "darkgrey"),
    legend.position = "bottom",  
    axis.text.x = element_text(size = 7),
    axis.text.y = element_text(size = 7)
  )
```

```{r, echo=FALSE}
with(soc_ins, {
  list("Young Unemployment" = lm_robust(formula = Income_inequality ~ 
                                          wdi_unempymilo)) |>
    msummary(coef_omit = "factor", stars = TRUE)
})
```

```{r, echo=FALSE}
## Basic Services on Economic Inequality ------------
### plotting the linear regression
soc_ins |> 
  filter(Year %in% 2000:2021) |>
  ggplot(aes(x = gov_size,
             y =  Income_inequality)) +
  geom_point() +
  geom_smooth(method = lm) +
  labs(title = "Basic Services on Economic Inequality (2000-2022)", 
       subtitle = "Income Inequality ~ Government Size", x = "Government Size", 
       y = "Income Inequality") +
  theme_minimal() +
  theme(
    axis.title = element_text(color = "darkgrey"),
    legend.position = "bottom",  
    axis.text.x = element_text(size = 7),
    axis.text.y = element_text(size = 7)
  )
```

```{r, echo=FALSE}
### linear regression 
with(soc_ins, {
  list("Government Size" = lm_robust(formula = Income_inequality ~ 
       gov_size
       )) |>
    msummary(coef_omit = "factor", stars = TRUE)
})
```

## Theory 3. Social inequality can predict social instability

### Correlation Plots

```{r, echo=FALSE}
soc_ins |> 
  dplyr::select(Country,
                "Political Terror" = pol_terr_scale_avg,
                "Post Elect Protest" = Riot_n_prot_after_elect,
                "Ed. Enroll. Tertiary" = sch_enrlmnt_tertiary_gross, 
                "Egalitarian Component Index" = egal_comp_idx, 
                Human_Inequality_Coefficient, 
                "Human Capital" = human_cap_idx,
                "Corruption" = control_of_corruption_estimate,
                Income_inequality,
                gov_size, 
                "Y. M. Unemployment" = wdi_unempymilo
                ) |>
  dplyr::select_if(is.numeric) |> 
  cor(use = "pairwise.complete.obs") |> 
  corrplot(method = "number", type = 'lower')
```

```{r, echo=FALSE}
soc_ins |> 
  dplyr::select(Country, 
                "Political Terror" = pol_terr_scale_avg,
                "Post Elect Protest" = Riot_n_prot_after_elect,
                "Ed. Enroll. Tertiary" = sch_enrlmnt_tertiary_gross, 
                "Egalitarian Component Index" = egal_comp_idx, 
                Human_Inequality_Coefficient, 
                "Human Capital" = human_cap_idx,
                "Corruption" = control_of_corruption_estimate,
                Income_inequality,
                gov_size, 
                "Y. M. Unemployment" = wdi_unempymilo
                ) |>
  dplyr::select_if(is.numeric) |> 
  cor(use = "pairwise.complete.obs") |>
  corrplot(type = 'upper', method = "shade") 
```

### Scatter Plots

```{r, echo=FALSE}

## Social Inequality on Social Instability ----------------------------

soc_ins |> 
  filter(Year %in% 2000:2022) |>
  ggplot(aes(x = sch_enrlmnt_tertiary_gross,
             y = pol_terr_scale_avg)) + 
  geom_jitter() +
  geom_smooth(method = lm) +
  labs(title = "Social Inequality on Social Instability (2000-2022)", 
       subtitle = "Political Terror ~ Educated Elites", x = "Educated Elites", 
       y = "Social Instability") +
  theme_minimal() +
  theme(
    axis.title = element_text(color = "darkgrey"),
    legend.position = "bottom",  
    axis.text.x = element_text(size = 7),
    axis.text.y = element_text(size = 7)
  )

soc_ins |> 
  filter(Year %in% 2000:2022) |>
  ggplot(aes(x = sch_enrlmnt_tertiary_gross,
             y = Riot_n_prot_after_elect)) + 
  geom_jitter() +
  geom_smooth(method = lm) +
  labs(title = "Social Inequality on Social Instability (2000-2022)", 
       subtitle = "Protests/Riots After Elections ~ Educated Elites", x = "Educated Elites", 
       y = "Social Instability") +
  theme_minimal() +
  theme(
    axis.title = element_text(color = "darkgrey"),
    legend.position = "bottom",  
    axis.text.x = element_text(size = 7),
    axis.text.y = element_text(size = 7)
  )
```

```{r, echo=FALSE}
### linear regression for social ineq on soc instability (polit_stab_absence_of_viol_terror) ----

with(soc_ins, {
  list(
    "School Enrollment Tertiary" = lm_robust(formula = pol_terr_scale_avg ~ 
       sch_enrlmnt_tertiary_gross + factor(Year) + factor(Country)),
    "School Enrollment Tertiary (Controls)" = lm_robust(formula = pol_terr_scale_avg ~
       sch_enrlmnt_tertiary_gross + wdi_unempymilo + control_of_corruption_estimate + 
       gov_size + Income_inequality + factor(Year) + factor(Country)),
    "Egalitarian component index" = lm_robust(formula = pol_terr_scale_avg ~ 
       egal_comp_idx + factor(Year) + factor(Country)),
    "Egalitarian component index (Controls)" = lm_robust(formula = pol_terr_scale_avg ~ 
       egal_comp_idx + wdi_unempymilo + control_of_corruption_estimate + 
       gov_size + Income_inequality + factor(Country) + factor(Year)),
    "Human Inequality Coefficient" = lm_robust(formula = pol_terr_scale_avg ~ 
       Human_Inequality_Coefficient + factor(Year) + factor(Country)),
    "Human Inequality Coefficient (Controls)" = lm_robust(formula = pol_terr_scale_avg ~ 
       Human_Inequality_Coefficient + wdi_unempymilo + control_of_corruption_estimate + 
       gov_size + Income_inequality + factor(Year) + factor(Country)),
    "Human Capital Index" = lm_robust(formula = pol_terr_scale_avg ~ 
       human_cap_idx + factor(Year) + factor(Country)),
    "Human Capital Index (Controls)" = lm_robust(formula = pol_terr_scale_avg ~ 
       human_cap_idx + wdi_unempymilo + control_of_corruption_estimate + 
       gov_size + Income_inequality + factor(Year) + factor(Country))
    ) |> 
    msummary(coef_omit = "factor", stars = TRUE)
  })

with(soc_ins, {
  list(
      "School Enrollment Tertiary" = lm_robust(formula = Riot_n_prot_after_elect ~ 
                                                 sch_enrlmnt_tertiary_gross + factor(Year) +
                                                 factor(Country)),
      "School Enrollment Tertiary (Controls)" = lm_robust(formula = Riot_n_prot_after_elect ~
                                                            sch_enrlmnt_tertiary_gross + wdi_unempymilo +
                                                            control_of_corruption_estimate + gov_size +
                                                            Income_inequality + factor(Year) + 
                                                            factor(Country)),
      "Egalitarian component index" = lm_robust(formula = Riot_n_prot_after_elect ~
                                                  egal_comp_idx + factor(Year) + factor(Country)),
      "Egalitarian component index (Controls)" = lm_robust(formula = Riot_n_prot_after_elect ~
                                                             egal_comp_idx + wdi_unempymilo +
                                                             control_of_corruption_estimate + 
                                                             gov_size + Income_inequality +
                                                             factor(Country) + factor(Year)),
      "Human Inequality Coefficient" = lm_robust(formula = Riot_n_prot_after_elect ~
                                                   Human_Inequality_Coefficient + factor(Year) + 
                                                   factor(Country)),
      "Human Inequality Coefficient (Controls)" = lm_robust(formula = Riot_n_prot_after_elect ~
                                                              Human_Inequality_Coefficient + wdi_unempymilo +
                                                              control_of_corruption_estimate + gov_size +
                                                              Income_inequality + factor(Year) + 
                                                              factor(Country)),
      "Human Capital Index" = lm_robust(formula = Riot_n_prot_after_elect ~ 
                                          human_cap_idx + factor(Year) + factor(Country)),
      "Human Capital Index (Controls)" = lm_robust(formula = Riot_n_prot_after_elect ~ 
                                                     human_cap_idx + wdi_unempymilo +
                                                     control_of_corruption_estimate + gov_size +
                                                     Income_inequality + factor(Year) + factor(Country))
    ) |> 
      msummary(coef_omit = "factor", stars = TRUE)
  })

```

```{r, echo=FALSE}
## Trust on Social Instability ----------------------------
### plotting the linear regression 
soc_ins |> 
  #filter(Year %in% 2000:2021) |>
  ggplot(aes(x = trust_in_people,
             y = Riot_n_prot_after_elect)) +
  geom_jitter() +
  geom_smooth(method = lm)
```

```{r, echo=FALSE}
### linear regression for trust on soc instability 
reg_tr_on_SI <- soc_ins |>
  #filter(Year %in% 2000:2021) |>
  lm_robust(formula = Riot_n_prot_after_elect ~ trust_in_people)
msummary(reg_tr_on_SI, stars = TRUE)

```

-   results

-   interpret

## Conclusion

Discussion "Those who don't know history are destined to repeat it." Though only a few know that the author of this aphorism was Spanish philosopher Jorge (George) Santayana, most of those who use it tacitly accept that history and, by extension, social dynamics may be governed by some kind of rules. Yet, the majority of historians emphatically dissent from such an idea. They reject the possibility that the dynamics of human societies be characterized by general regularities or governed by rules. For them, because a person's behavior is unpredictable, the dynamics of the societies they constitute cannot be predicted either. Such unpredictability would make any attempt to mathematically or statistically model it a gross oversimplification, unable to explain anything meaningful about social or historical dynamics--let alone predicting them.

This paper is the first step in a long term project that seeks to bridge both perspectives to understand what economic, political, social, environmental, and international factors may lead to social instability (discontent, unrest, and violence). The project aims at devising a model capable of predicting when unrest and violence are likely to develop so policymakers can make adjustments avoid discontent, ensure fairness, and safeguard political and economic stability. It also aims at becoming a risk analysis tool that fed with up-to-date intelligence help investors to plan business and take precautions accordingly, and travelers/visitors to avoid distressing and risky situations.

Compare to others





Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
